---
output: html_document
---

```{r setup, include=FALSE}
source("style.R")
```


# A very short introduction to R

R is a large and complex topic.
Even those who use it every day only know a small subset of what there is to know.
However, you dont need to know very much to do some very impressive things.
This section is a **very** short introduction to the most basic aspects of R to get you started.
Look at the list of resources and the end of this section for more complete tutorials.

## What is R?

Unlike spreadsheet programs, like excel, R is a text-based "interactive" program, meaning that you use it by typing commands.
If you type a valid command, R will do something, like printing something to the screen or modifying a peice of data.
Where you type commands is referred to as the "R console".
If you are using RStudio, this is the lower left window.
Here is what an R console looks like when R is first started:

```
R version 3.4.4 (2018-03-15) -- "Someone to Lean On"
Copyright (C) 2018 The R Foundation for Statistical Computing
Platform: x86_64-pc-linux-gnu (64-bit)

R is free software and comes with ABSOLUTELY NO WARRANTY.
You are welcome to redistribute it under certain conditions.
Type 'license()' or 'licence()' for distribution details.

  Natural language support but running in an English locale

R is a collaborative project with many contributors.
Type 'contributors()' for more information and
'citation()' on how to cite R or R packages in publications.

Type 'demo()' for some demos, 'help()' for on-line help, or
'help.start()' for an HTML browser interface to help.
Type 'q()' to quit R.

> |
```

For example, if I type `1 + 1` into the console, R will do the math and "return" the result. 


```
> 1+1
[1] 2
```

The thing after the `>` is what I typed and the line below is what was returned.

Since I did not save the result in a variable, it was printed instead.
The `[1]` at the start keeps track of how many values have been printed.
Its not too useful when their is only a single value, as in this case, but gets more useful when many values are printed.
For example, I could make a `numeric` with 100 values from 1 to 100 by typing:

```
> 1:100
  [1]   1   2   3   4   5   6   7   8   9  10  11  12  13  14  15  16  17  18  19  20  21  22  23  24  25  26  27  28  29  30  31
 [32]  32  33  34  35  36  37  38  39  40  41  42  43  44  45  46  47  48  49  50  51  52  53  54  55  56  57  58  59  60  61  62
 [63]  63  64  65  66  67  68  69  70  71  72  73  74  75  76  77  78  79  80  81  82  83  84  85  86  87  88  89  90  91  92  93
 [94]  94  95  96  97  98  99 100
```

Now the `[32]` on the second line means that `32` is the 32nd element.
Every type of variable will have a different way of being printed.
In this case, the variable is a `numeric` type.

## Variables

In programmer-speak, any peice of data is called a "variable". 
There are different types (or "classes") of variables such as `character` (text), `numeric` (numbers), and `data.frame`s (tables).
You can "save" the result of commands in R using `<-` like so:

```
> a_sequence <- 1:100
```

Note how nothing is printed since the result of `1:100` was saved in the `a_sequence` variable, but it can still be printed by entering the name of the variable:

```
> a_sequence
  [1]   1   2   3   4   5   6   7   8   9  10  11  12  13  14  15  16  17  18  19  20  21  22  23  24  25  26  27  28  29  30  31
 [32]  32  33  34  35  36  37  38  39  40  41  42  43  44  45  46  47  48  49  50  51  52  53  54  55  56  57  58  59  60  61  62
 [63]  63  64  65  66  67  68  69  70  71  72  73  74  75  76  77  78  79  80  81  82  83  84  85  86  87  88  89  90  91  92  93
 [94]  94  95  96  97  98  99 100
```

Now, anythime `a_sequence` is used it would be as if `1:100` was typed instead (although the calculation would not be done again).


## Functions

Any command/tool/action in R is a function.
Most functions are used by putting the input to the function (if any) in parentheses following the function name.
For example, the `length` function returns how many things are in a variable:

```
> length(a_sequence)
[1] 100
```

Even things like `+` and `:` are functions, although these are used in a special way.
For example, they could also be used this way, just to demonstrate the idea:

```
> `:`(1, 10)
 [1]  1  2  3  4  5  6  7  8  9 10
> `+`(1, 1)
[1] 2
```

The vast majority of functions are used like `length`.
Perhaps confusingly, functions are also varaibles and can be created and "saved".
For example, the following code creates a function to add two numbers together and saves in the variable `add`: 

```
> add <- function(x, y) { x + y }
```

We can now use this like any other function: 

```
> add(1, 1)
[1] 2
```

And it can be printed like any other variable:

```
> add
function(x, y) { x + y }
```

Functions can range in complexity from simple ones like `add` to very long and complex functions that call other custom functions.


## R packages

An R pacakge is a set of user-defined functions organized so that people can easily share and use them.
Most of the functions used by most R users are from R packages rather than those supplied by "base" R.
R packages can be installed in a few ways, but the most common is to download them from The Comprehensive R Archive Network (CRAN) using the `install.packages` function.
For example `stringr` is an R package that supplies functions to work with text.

```
> install.packages("stringr")
```

Once installed, a package must be "loaded" using the `library` function before any functions it supplies can be used:

```
> library("stringr")
```

Now we can use functions from the `stringr` package.
For example, the `str_count` function counts the number of times a peice of text occurs in a larger peice of text:

```
> str_count(string = "R is awesome!!!", pattern = "!")
[1] 3
```

## Getting help/documentation

One of the most important skills in R is looking up (and interpreting) the built-in documentation.
You can get the help for **any** R function by prefixing it with a question mark like so: 

```{r, eval=FALSE}
?getwd
```

Help files have an overview of:

 - purpose of a function
 - parameters it takes
 - output it yields
 - examples demonstrating its usage
 
This built-in documentation for each function can be a bit terse and hard to interpret and its quality varies greatly between packages.
R pacakges will also usually have "vignettes", which combine examples and explaination for how functions in a pacakge are generally used.
These can be found online or accessed in the installed pacakge using the `browseVignettes` function:

```{r eval=FALSE}
browseVignettes(package = "metacoder")
```

Finally, some of the best documentation can be found online.
For some of the more popular pacakges, like `dplyr` and `ggplot2`, the best documentation is from unofficial sources, like blogs, books, and question/answer sites like [Stack Overflow](https://stackoverflow.com/).


## What to do when you dont know what to do

When you run into problems or dont know how to do something, **it will save you a lot of time** if you do an internet search describing the problem you have before trying to fix it by guessing, reading through documenttation, or making a custom solution.
If the problem has an error message, then **copy and paste the error into a search engine**. 
At least 95% of problems are well documented and discussed.
If you think "there must be a better way to do this", there almost certantly is and some nice person on the internet spent hours writing a blog post about it or ansewering a question on [Stack Overflow](https://stackoverflow.com/).
The trick is knowing what to type since the problem/concept might take some jargon you dont know to describe.
Try describing your problem a few ways before giving up.
If you know someone who uses R more, try asking them what to search for rather than asking them to help you themselves (unless they are super nice and/or have nothing better to do).
For example, "I want to combine two tables based on the content of a column they both share" is a very common need, but it might be hard to get Google to undertand what you want.
Someone who uses R a lot will tell you to search for "joining" or "joins", which will lead you in the right direction, and will take only a few seconds of their time.

## More resources

This was only the most basic concepts in R, but there are lots of free tutorials and info online to learn more.
Searching "beginner R tutorial" on the internet will show hundreds of free resources.
Here are a few resources for beginners to learn R:

* [Jenny Bryan's 545 statistics in R course content](http://stat545.com/topics.html)
* [Data Carpentry: R for data analysis and visualization of Ecological Data](http://www.datacarpentry.org/R-ecology-lesson/index.html)
* [Software Carpentry: Programming with R](http://swcarpentry.github.io/r-novice-inflammation/)
* [Software Carpentry: R for Reproducible Scientific Analysis](http://swcarpentry.github.io/r-novice-gapminder/)
* [DataCamp's free introduction to R course](https://www.datacamp.com/courses/free-introduction-to-r)
