---
output: html_document
bibliography: "bibtexlib.bib"
---

```{r setup, include=FALSE}
source("style.R")
```

# Installing required software and data

## Installing R

R is a computer language that focuses on statistics, data science, and visualization.
It can be installed on all common operating systems.
R can be downloaded here:

https://cran.r-project.org/


## Installing RStudio

RStudio is an "Integrated Development Environment" (IDE) which is a fancy way of saying "text editor with extra programming-related tools".
RStudio makes it easier to use R, but is not needed to use R.
Most R users seem to use RStudio, so we are recommending it for this workshop.
The main version of RStudio, which is free and open source, can be downloaded here:

https://www.rstudio.com/products/rstudio/download/#download


## Installing the required R packages

In addition to tools included with every R installation, it is common to install "R packages", which are sets of tools bundled together in a standardize way, making them easy to install.
In fact, most functionality used in R is from R packages, not "base" R. 
The following packages are utilized in this primer.
Clicking of their name will lead to documentation (perhaps unofficial) for the package.

1. [vegan](http://cc.oulu.fi/~jarioksa/opetus/metodi/vegantutor.pdf): Implements many standard statistical techniques used in ecological research. Is used extensively by packages like `phyloseq` and `metacoder`. [@dixon2003vegan]
1. [metacoder](https://github.com/grunwaldlab/metacoder#an-r-package-for-metabarcoding-research-planning-and-analysis): Manipulation and visualization of taxonomic data, particularly those from aplicon metagenomics research. [@foster2017metacoder]
1. [phyloseq](https://joey711.github.io/phyloseq/index.html): Tools and visualization for microbiome data specifically. [@mcmurdie2013phyloseq]
1. [ggplot2](http://r-statistics.co/Complete-Ggplot2-Tutorial-Part1-With-R-Code.html): Awesome graphing package. [@wickham2009ggplot2]
1. [dplyr](http://stat545.com/block009_dplyr-intro.html): A package for manipulating tabular data with a cohesive and intuitive set of commands. A popular alternative to base R methods. [@wickham2015dplyr]
1. [readr](https://cran.r-project.org/web/packages/readr/vignettes/readr.html): Makes reading tabular data from files easier.

More R resources useful for microbiome data analysis can be found here:

https://microsud.github.io/Tools-Microbiome-Anlaysis/ 

You can enter the following script in the R console to install these packages:

```{r, eval = FALSE}
# Install phyloseq from Bioconductor
source('http://bioconductor.org/biocLite.R')
biocLite("phyloseq")

# Install the rest of the packages from CRAN
install.packages(c("vegan", "metacoder", "ggplot2", "dplyr", "readr"),
                 repos = "http://cran.rstudio.com",
                 dependencies = TRUE)
```

We wrote and actively maintain *metacoder* `r citep(bib["foster2017metacoder"])` and it
is heavily relied upon in this primer. You can think of an R package as a library of 
functions written and curated by someone in the R user
community, which you can be loaded into R for use.

Congratulations! You should now be all set for using R. Loading data and
conducting your first analysis will be the topic of the next chapter. But before
we go there lets provide a few useful resources.


## References
