---
title: "Diversity statistics"
output: html_document
bibliography: "bibtexlib.bib"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load example data

If you are starting the workshop at this section, or had problems running code in a previous section, use the following to load the data used in this section.
If `obj` and `sample_data` are already in your environment, you can ignore this and proceed.

```{r}
load("clean_data.Rdata")
```

## Measures of diversity

Diversity in the ecological sense is intuitivly understood as the complexity of a community of organisms.
There are many ways to quantify this complexity so that we can compare communities objectivly.
The two main categories of methods are known as **alpha diversity** and **beta diversity** [@whittaker1960vegetation].
Alpha diversity measures the diversity within a single sample and is generally based on the number of realtive abundance of species.
Beta diversity also uses the number of realtive abundance of species, but measures variation between samples.

```{r eval = FALSE}
library(magrittr)
library(AHMbook)
# Simulate true community
comm_stats <- simComm(type = "counts", nsite = 2, nspec = 100, nrep = 5, show.plot = FALSE)
comm_counts <- as.data.frame(t(comm_stats$N)) + 1
table(comm_counts$site1)

hist(comm_counts$site1)
hist(comm_counts$site2)

# Take samples
rep(rownames(comm_counts), comm_counts$site1)
```


## Alpha (within sample) diversity 

## Beta (between sample) diversity


## References

```{r, child="_sessioninfo.Rmd"}
```